Machine Learning Tools for Open Cluster Characterization with Gaia DR2 Data

### 2021-02-21 paper/paper.tex
paper: Fix NGC 2682 Clusterix method name
Add final corrections
paper: Final changes
Co-Authored-By: Cesar Guzmann <15159632+cguz@users.noreply.github.com>

\resumen{
The characterization and understanding of \emph{Open Clusters} (OCs) allow us to
understand better properties and mechanisms about the Universe such as stellar
formation and the regions where these events occur. They also provide information
about stellar processes and the evolution of the galactic disk.

@@

The characterization of an OC refers to the statistical determination
of the stars that form it. Therefore, the characterization is based on
the probability of a star to be a member of the cluster~\cite{sampedro2016caracterizacion}.

@@

In the present paper, we propose a machine learning model capable of characterizing
OCs inside a stellar region with no previous knowledge about the region.
Our model takes advantage of some features (such as proper motion in right ascension
and declination, and parallax) to train an Artificial Neural Network.
The ANN clusters all the stars within a given region in groups.
One of those groups will be the open cluster that interests us.

@@

\section{Related work}
\label{sec:related_works}

@@

Clusterix critically depends on the selection of three radii in the studied region.
The inner radius contains stars belonging or not to the cluster. While the outermost
radii defines a ring which only contains stars that do not belong to the cluster.
In addition, it is also sensitive to other parameters, such as the soften parameter
\emph{h}, or other restrictions related with the searched proper motions.

If the analysis performed by Clusterix is successful, it returns the probability of
each star to belong to the open cluster. The results can be imported then into TOPCAT
to continue the refining process to get a valid characterization.

Another recent approach~\cite{castro2020hunting} solves the problem using machine
learning techniques. It searches systematically for overdensities in the astrometric
space of the galactic disk and a subsequent identification of open clusters using
photometric information. It includes two phases:

\begin{enumerate}
  \item it employs DBSCAN, an unsupervised clustering algorithm, to search for
        overdensities,
  \item and it applies a deep learning ANN to identify isochrone patterns within
        the detected overdensities and thus proceed to confirm them as OCs. They
        trained previously the ANN with magnitude diagrams.
\end{enumerate}

They execute their experiments in the Barcelona Supercomputing Center, MareNostrum 42.
Thereby, the neural network handles the image recognition process with isochrone
patterns without applying theoretical models derived from values such as metallicity
or masses, among other. Their work concludes with the identification of 582 new open
clusters distributed along the galactic disk for a galactic declination bellow 20
degrees, increasing the number of known open clusters by 45\%.

In contrast to~\cite{castro2020hunting}, we lack of a super computer like MareNostrum
for our experiments. For this reason, our aim is to obtain a new novel method that
allows the characterization of open clusters without doing a blind search for clusters
characterization. We advocate of the idea of building an unsupervised and
non-parameterized method. So, it can be suitable for automated processes.

@@

In the literature, we have several clustering algorithms, such as K-Means, Mean-Shift
Clustering, DBSCAN~\cite{ester1996density} among other, that can help us to achieve
our aim. Each one behaves better according to the distribution of the objects to
clusterize. However, we need to set a large number of clusters to find the open cluster
we are looking for. It complicates the identification of the OC because most of the
time falls in many outliers.

@@

With all this in mind, our method is based on DEC model. We adapted it to manage data
groups based on the dynamic properties of the stars. Thus, we have an unsupervised
clustering model for open cluster characterization. The model fits a wide range of
clusters without the need for fine-tuning a high number of hyperparameters.

\subsection{Selection of the catalogue}

In this work we make use of Gaia DR2 since DR3 has not been released in time for us to
include it. Gaia DR2 is a multidimensional dataset obtained by ESA's Gaia mission
(located at L2, 1.5 million kilometers from Earth) and operational since 2014. The
catalogue has high precision and accuracy astrometric data for more than 1.7 billion
stellar sources, and magnitudes in three photometric filters (G, BP and RP) for more
than 1,300 million sources.

We start by selecting a region from the OpenClust~\cite{dias2002new} catalogue and
we downloaded it from Gaia DR2 database. The radius of the downloaded region from Gaia
is 1.5 times larger than the one registered in OpenClust, ensuring to include several
stars that do not belong to the open cluster.

@@

However, we are not going to use these raw features. Instead, we are taking a combination
of them. Figure~\ref{fig:features_melotte_22} shows a pairwise relationship of our
combination of features from Melotte 22.

@@

\subsection{Deep Open Clustering of stars}
\label{sec:deep_open_clustering}

In this section we explain in more detail our model.

In order to define our model, one of the main problems is that we do not have a labeled
dataset to train a supervised model. Thus, we have to deal with an unsupervised self-trained
model. We adapted the Unsupervised \emph{DEC} model~\cite{xie2016unsupervised} to our requirements.

Same as the DEC model, we have two main components:

\begin{itemize}
  \item \textbf{deep autoencoder:} It is trained before passing the data through the
        \emph{clustering layer}. It is composed by encoder layers followed by decoder
        layers. Recent research has shown that this autoencoder provides meaningful and
        well-separated representations on real-world datasets~\cite{vincent2010stacked}
        and~\cite{hinton2006reducing} to the DEC model.
  \item \textbf{clustering layer:} This layer recibes data transformed by the autoencoder
        and it is iteratively trained until a convergence criterion is met.
\end{itemize}

@@

As it was described in Section~\ref{sec:feature_selection}, the number of features we
deal is not too large. This latent space helps us to start in a reduced number of
features and avoids the \emph{``curse of dimensionality``}~\cite{bellman1961curse}.

The autoencoder is pretrained before fitting the model to generate predictions. Then,
the encoder layers of the autoencoder are used with the aim of transforming input data to
the latent space \(Z\). Once the data has been transformed, a K-Means clusterer is used in
order to make an initial clustering. K-Means cluster centers are used as the initial
weights for the clustering layer.

With that initial configuration, the model iterates alternating between computing an auxiliary
target distribution (Soft Assignment) and minimizing the Kullback-Leiber (KL)
divergence~\cite{kullback1951information} to it.
This unsupervised algorithm allows us to improve the clustering.

@@

In the soft assignment stage,
the \emph{Student's t-distribution} is used as a kernel to measure the similarity
between the embedded points and the cluster centroid.
While in the KL divergence minimization the algorithm iteratively refines clusters by learning
from their high confidence assignments with the help of an auxiliary target distribution.
The model is trained by matching the soft assignment to the target distribution.
The choice of this target distribution is crucial for DEC's performance.
In this work we have taken the target distribution from DEC's original paper~\cite{xie2016unsupervised},
which is defined in Equation~\ref{eq:student_tdistribution}.

\begin{figure*}[!hbt]
  \centering
  \includegraphics[scale=0.8]{../figures/dec_diagram.pdf}
  \caption{DEC model layer setup}
  \label{fig:dec_model_setup}
\end{figure*}

Figure~\ref{fig:dec_model_setup} shows the layer setup of our DEC model.
It is simpler than the one tested on the original paper~\cite{xie2016unsupervised},
since the number of selected features in our work is smaller than in the original one.
Therefore, using the same configuration would result in a model so
powerful that would incur in overfitting issues unable to make right predictions.

@@

Finally, we can refine this selection by filtering those stars which are bellow and above
the 0.10 and 0.90 quantiles for each group, respectively. That way we remove the most
doubtful values from the selection.

@@

Figure~\ref{fig:kmeans_comparisons_melotte_22} shows an example. It describes the results of
applying K-Means with N=2, N=5 and N=8 in Melotte 22. As we can see, large values of the number
of clusters allow us to isolate more accurately the resonance in parallax at \(\approx 7.3mas\).

The disadvantage is that it forms more groups of stars, which complicates the task of finding the
desired OC, since we would like to get just two groups. Therefore, with a K-Means we have to find
a way to set the right value for the number of clusters to isolate the searched cluster without
creating too many groups. To solve this issue, we can try to estimate the best number of clusters
by using the \emph{silhouette score}~\cite{rousseeuw1987silhouettes}.

@@

\begin{itemize}
  \item NGC 2516 (Ap.~\ref{sec:ngc2516}) has it proper motion center deviated from the origin but
        it is embedded inside a big cloud of stars with similar proper motions.
  \item NGC 2632 (Ap.~\ref{sec:ngc2632}), in addition to Melotte 22, ia s cluster whose proper motion center
        is not located at the origin and has a well separated parallax center.
  \item NGC 2682 (Ap.~\ref{sec:ngc2682}), on the other hand, has its parallax centered within the region's gaussian,
        which complicates its detection although its proper motion center is deviated from the origin.
  \item Melotte 22 (Ap.~\ref{sec:melotte22}), as well as Melotte 25, is a cluster closer to us than the other.
        That makes its membership stars to be more scattered than previous clusters which are more compact.
\end{itemize}

@@

\subsection{Future work}

@@

\begin{table}[htbp]
  \begin{center}
    \resizebox{\columnwidth}{!}{
      \begin{tabular}{l|c|c|c|c}
        \textbf{Method} & \emph{\(\mu_{\alpha}\) \((mas \cdot yr^{-1})\)} & \emph{\(\mu_{\delta}\) \((mas \cdot yr^{-1})\)}
        & \emph{\( \varpi \) \((mas)\)} & \emph{\# stars} \\
        \hline
        \textbf{Simbad} & -10.9737 \( \pm \) 0.0064 & -2.9396 \( \pm \) 0.0063 & 1.1325 \( \pm \) 0.0011 & 1194 \\
        Clusterix+TOPCAT & -10.970 \( \pm \) 0.322 & -2.958 \( \pm \) 0.327 & 1.142 \( \pm \) 0.080 & 649 \\
        K-Means & -8.616 \( \pm \) 0.15 & -3.710 \( \pm \) 0.16 & 1.196 \( \pm \) 0.01 & 1374 \\
        DEC & -8.926 \( \pm \) 0.15 & -3.550 \( \pm \) 0.15 & 1.144 \( \pm \) 0.005 & 1238 \\
        \textbf{DEC (filt.)} & -9.619 \( \pm \) 0.13 & -3.317 \( \pm \) 0.13 & 1.140 \( \pm \) 0.003 & 990 \\
      \end{tabular}
    }
    \caption{NGC 2682 results.}
    \label{tab:app_results_ngc_2682}
  \end{center}
\end{table}

>>>>

\resumen{
Characterize and understand \emph{Open Clusters} (OCs) allow us to
understand better properties and mechanisms about the Universe such as stellar
formation and the regions where these events occur. They also provide information
about stellar processes and the evolution of the galactic disk.

@@

Characterize an OC refers to the statistical determination of the stars that
form it. Therefore, the characterization is based on the probability of a star
being a member of the cluster~\cite{sampedro2016caracterizacion}.

@@

In the present paper, we propose a machine learning model capable of characterizing
OCs inside a stellar region with no previous knowledge about the region.
Our model takes advantage of some features (such as proper motion in right ascension
and declination, and parallax) to train an ANN.
The ANN clusters all the stars within a given region in groups.
One of those groups will be the open cluster that interests us.

@@

\section{Related Works}
\label{sec:related_works}

@@

Clusterix depends fundamentally on the selection of three radii in the studied region.
The inner radius contains stars that belong or not to the cluster. While the outermost
radii define a ring that only contains stars that do not belong to the cluster.
Furthermore, it is also sensitive to other parameters, such as the soften parameter
\(h\), or other restrictions related to the searched proper motions.

If the analysis performed by Clusterix is successful, it returns the probability that
each star belongs to the open cluster. The results can then be imported into TOPCAT
to continue the refinement process and obtain a valid characterization.

Another recent approach~\cite{castro2020hunting} solves the problem using machine
learning techniques. It searches systematically for overdensities in the astrometric
space of the galactic disk and subsequent identification of open clusters using
photometric information. It includes two phases:

\begin{enumerate}
  \item it employs DBSCAN, an unsupervised clustering algorithm, to search for
        overdensities,
  \item and applies a deep learning ANN to identify isochrone patterns within
        the detected overdensities and thus proceed to confirm them as OCs. They
        previously trained the ANN with magnitude diagrams.
\end{enumerate}

They carry out their experiments at the Barcelona Supercomputing Center, MareNostrum 42.
Thereby, the neural network handles the image recognition process with isochrone
patterns without applying theoretical models derived from values such as metallicity
or masses, among others. Their work concludes with the identification of 582 new open
clusters distributed along with the galactic disk for a galactic declination below 20
degrees, increasing the number of known open clusters by 45\%.

In contrast to~\cite{castro2020hunting}, we lack a supercomputer like MareNostrum
for our experiments. For this reason, our goal is to obtain a new novel method that
allows the characterization of open clusters without doing a blind search for
characterization clusters. We advocate the idea of building an unsupervised and
non-parameterized method. Therefore, it can be suitable for automated processes.

@@

In the literature, we have several clustering algorithms, such as K-Means, Mean-Shift
Clustering, DBSCAN~\cite{ester1996density} among others, that can help us achieve
our aim. Each one behaves better according to the distribution of the objects to
be grouped. However, we need to establish a large number of clusters to find the open cluster
we are looking for. It complicates the identification of the OC because most of the
time falls in many outliers.

@@

With all this in mind, our method is based on the DEC model. We adapted it to manage data
groups based on the dynamic properties of the stars. Thus, we have an unsupervised
clustering model for open cluster characterization. The model fits a wide range of
clusters without the need for fine-tuning a high number of hyperparameters.

\subsection{Selection of the catalog}

In this work, we make use of Gaia DR2 as DR3 has not been released in time for us to
include it. Gaia DR2 is a multidimensional dataset obtained by ESA's Gaia mission
(located at L2, 1.5 million kilometers from Earth) and operational since 2014. The
catalog has high precision and accuracy astrometric data for more than 1.7 billion
stellar sources and magnitudes in three photometric filters (G, BP, and RP) for more
than 1,300 million sources.

We start by selecting a region from the OpenClust~\cite{dias2002new} catalog and
downloading it from the Gaia DR2 database. The radius of the downloaded region from Gaia
is 1.5 times greater than that recorded in OpenClust, ensuring that it includes several
stars that do not belong to the open cluster.

@@

However, we are not going to use these raw features. Instead, we take a combination
of them. Figure~\ref{fig:features_melotte_22} shows a pairwise relationship of our
combination of features from Melotte 22.

@@

\subsection{Deep Open Clustering of Stars}
\label{sec:deep_open_clustering}

In this section, we explain our model in more detail.

One of the main problems to define our model is that we do not have a labeled
dataset to train a supervised model. To solve it, we employ an unsupervised self-trained
model. Thus, we adapted the Unsupervised \emph{DEC} model~\cite{xie2016unsupervised}
to our requirements.

Like the DEC model, we have two main components:

\begin{itemize}
  \item \textbf{deep autoencoder:} it is a neural network that learns to copy its input
        to its output. It is composed of encoder layers followed by decoder layers.
        We train it before passing the data through the next component, the
        \emph{clustering layer}. Recent research~\cite{vincent2010stacked, hinton2006reducing}
         showed that the autoencoder provides meaningful and well-separated representations
         on real-world datasets.
  \item \textbf{clustering layer:} it receives data transformed by the autoencoder
        and iteratively is trained until a convergence criterion is met.
\end{itemize}

@@

We trained the autoencoder before fitting the model to generate predictions.
Then, we use the encoder layers of the autoencoder with the aim of transforming
input data to the latent space \(Z\). Once the data has been transformed, a K-Means
clusterer is used to make an initial clustering. K-Means cluster centers are used
as the initial weights for the clustering layer.

With that initial configuration, the model iterates alternating between computing an
auxiliary target distribution (soft assignment) and minimizing the Kullback-Leiber (KL)
divergence~\cite{kullback1951information} to it. This unsupervised algorithm allows us
to improve the clustering.

As it was described in Section~\ref{sec:feature_selection}, the number of features we
deal is not too large. However, this latent space helps us to start with a reduced number
of features and avoids the \emph{``curse of dimensionality''}~\cite{bellman1961curse}.

In the soft assignment stage, we used \emph{Student's t-distribution} as a kernel to
measure the similarity between the embedded points and the cluster centroid.
While in the KL divergence minimization the algorithm iteratively refines clusters by learning
from their high confidence assignments with the help of an auxiliary target distribution.
The model is trained by matching the soft assignment to the target distribution.
The choice of this target distribution is crucial for model's performance.

In this work, we took the target distribution~\cite{xie2016unsupervised}
defined in Equation~\ref{eq:student_tdistribution}.

@@

This target distribution has the following properties:

\begin{itemize}
  \item strengthen predictions (i.e., improve cluster purity),
  \item put more emphasis on data points assigned with high confidence,
  \item normalize loss contribution of each centroid to prevent large clusters
        from distorting the hidden feature space.
\end{itemize}

\begin{figure*}[!hbt]
  \centering
  \includegraphics[scale=0.8]{../figures/dec_diagram.pdf}
  \caption{Layer setup of our model}
  \label{fig:dec_model_setup}
\end{figure*}

Figure~\ref{fig:dec_model_setup} shows the layer setup of our model.
It is simpler than the used in~\cite{xie2016unsupervised} because we selected a smaller
number of features. Therefore, using the same configuration would result in a model so
powerful that it would incur overfitting issues that cannot make the correct predictions.

@@

Finally, we can refine this selection by filtering those stars that are below and above
the 0.10 and 0.90 quantiles for each group, respectively. That way we remove the most
doubtful values from the selection.

@@

Figure~\ref{fig:kmeans_comparisons_melotte_22} shows an example. It describes the results of
applying different configurations of K-Means in Melotte 22, N=2, N=5 and N=8. For each configuration
of K-Means, we show the proper motion in right ascension and declination (figures in the top position),
and the parallax distribution (figures in the down position).

As we can see in the parallax distribution, lower values of the number of clusters (e.g., N=2) assign
many stars to the group \emph{0} (purple) that should belong to group \emph{g1}. These badly assigned stars
are the ones that have the parallax far from the 7.3mas resonance. In contrast, large values of the
number of clusters allow us to isolate more accurately the resonance in parallax at \(\approx 7.3mas\).
For instance, at N=8, more groups are better compartmentalized. Group \emph{g2} would be the one corresponding
to the OC that we are looking for, and all the other groups are those that we would like to be a single
group. That is the stars that do not belong to the OC. The disadvantage is that K-Means with N=8 forms more
groups of stars, which complicates the task of finding the desired OC, since we would like to get just two
groups. Therefore, with a K-Means we have to find  a way to set the right value for the number of clusters
to isolate the searched cluster without creating too many groups. To solve this issue, we can try to
estimate the best number of clusters by using the \emph{silhouette score}~\cite{rousseeuw1987silhouettes}.

@@

\begin{itemize}
  \item NGC 2516 (Ap.~\ref{sec:ngc2516}) has it proper motion center deviated from the origin but
        it is embedded inside a big cloud of stars with similar proper motions.
  \item NGC 2632 (Ap.~\ref{sec:ngc2632}), in addition to Melotte 22, is a cluster whose proper motion center
        is not located at the origin and has a well separated parallax center.
  \item NGC 2682 (Ap.~\ref{sec:ngc2682}), on the other hand, has its parallax centered within the region's gaussian,
        which complicates its detection although its proper motion center is deviated from the origin.
  \item Melotte 22 (Ap.~\ref{sec:melotte22}), as well as Melotte 25, is a cluster closer to us than the other.
        That makes its membership stars to be more scattered than previous clusters which are more compact.
\end{itemize}

@@

\subsection{Future Works}

@@

\begin{table}[htbp]
  \begin{center}
    \resizebox{\columnwidth}{!}{
      \begin{tabular}{l|c|c|c|c}
        \textbf{Method} & \emph{\(\mu_{\alpha}\) \((mas \cdot yr^{-1})\)} & \emph{\(\mu_{\delta}\) \((mas \cdot yr^{-1})\)}
        & \emph{\( \varpi \) \((mas)\)} & \emph{\# stars} \\
        \hline
        \textbf{Simbad} & -10.9737 \( \pm \) 0.0064 & -2.9396 \( \pm \) 0.0063 & 1.1325 \( \pm \) 0.0011 & 1194 \\
        Clusterix & -10.970 \( \pm \) 0.322 & -2.958 \( \pm \) 0.327 & 1.142 \( \pm \) 0.080 & 649 \\
        K-Means & -8.616 \( \pm \) 0.15 & -3.710 \( \pm \) 0.16 & 1.196 \( \pm \) 0.01 & 1374 \\
        DEC & -8.926 \( \pm \) 0.15 & -3.550 \( \pm \) 0.15 & 1.144 \( \pm \) 0.005 & 1238 \\
        \textbf{DEC (filt.)} & -9.619 \( \pm \) 0.13 & -3.317 \( \pm \) 0.13 & 1.140 \( \pm \) 0.003 & 990 \\
      \end{tabular}
    }
    \caption{NGC 2682 results.}
    \label{tab:app_results_ngc_2682}
  \end{center}
\end{table}
